---
title: "Getting Started with fmriAR"
author: "fmriAR Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with fmriAR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

`fmriAR` provides lightweight tools for estimating autoregressive (AR) or autoregressive–moving-average (ARMA) noise models from fMRI residuals and applying matched prewhitening to both design matrices and voxel-wise data. The C++ core keeps large-scale whitening fast, while the R interface stays compact and flexible.

This vignette walks through a typical workflow. We start by fitting an AR model to residuals from an initial OLS regression, then apply the resulting run-aware whitening plan to both the design matrix and the voxel data. With the innovations in hand we check that low-lag autocorrelation has been suppressed. We then cover censored (scrubbed) frames, correcting the bias that the GLM projection puts into residual autocorrelation, parcel pooling with multiscale shrinkage, ARMA models with automatic order selection, AFNI-style restricted models, and standard errors.

## Simulating data for the examples

To make the examples reproducible, we simulate a modest fMRI-style dataset with two runs, a small design matrix, and voxel-specific AR(2) noise.

```{r simulate}
library(fmriAR)
set.seed(42)

n_time <- 240
n_vox <- 60
runs <- rep(1:2, each = n_time / 2)

# Design matrix: intercept + task regressor
X <- cbind(
  intercept = 1,
  task = rep(c(rep(0, 30), rep(1, 30)), length.out = n_time)
)

# Generate voxel time-series with AR(2) noise
phi_true <- matrix(0, nrow = n_vox, ncol = 2)
phi_true[, 1] <- 0.5 + rnorm(n_vox, sd = 0.05)
phi_true[, 2] <- -0.2 + rnorm(n_vox, sd = 0.05)

Y <- matrix(0, n_time, n_vox)
innov <- matrix(rnorm(n_time * n_vox, sd = 1), n_time, n_vox)
for (v in seq_len(n_vox)) {
  for (t in seq_len(n_time)) {
    ar_part <- 0
    if (t > 1) ar_part <- ar_part + phi_true[v, 1] * Y[t - 1, v]
    if (t > 2) ar_part <- ar_part + phi_true[v, 2] * Y[t - 2, v]
    Y[t, v] <- ar_part + innov[t, v]
  }
}

# Add task signal to half the voxels
beta <- c(0, 1.5)
Y_signal <- drop(X %*% beta)
Y[, 1:30] <- sweep(Y[, 1:30], 1, Y_signal, "+")

# Residuals from initial OLS fit
coeff_ols <- qr.solve(X, Y)
resid <- Y - X %*% coeff_ols
```

## Fitting an AR model

The primary entry point is `fit_noise()`. When `method = "ar"` and `p = "auto"`, the function selects the AR order per run using BIC (searching up to `p_max`), then enforces stationarity. An explicit order such as `p = 2` is used as given.

`exact_first = "ar1"` (the default; the name is historical) whitens the start of every run, and every segment after a censored frame, with the exact stationary initialisation of the fitted model. For any AR or ARMA order this makes the whitening identical to exact GLS within each segment; for an AR(1) it reduces to the familiar `sqrt(1 - phi^2)` scaling of the first sample.

```{r fit-ar}
plan_ar <- fit_noise(
  resid,
  runs = runs,
  method = "ar",
  p = "auto",
  p_max = 6,
  exact_first = "ar1",
  pooling = "run"
)

plan_ar
```

The returned `fmriAR_plan` holds run-specific AR coefficients and metadata about the estimation procedure.

## Whitening design and data matrices

`whiten_apply()` takes the plan, design matrix, and observed data and returns whitened versions that can be used in GLS estimation or downstream analyses.

```{r whiten-run}
whitened <- whiten_apply(plan_ar, X, Y, runs = runs)
str(whitened)
```

By default the function returns whitened `X` and `Y`; your inputs are never modified. Whitening runs in parallel over voxels when the package was built with OpenMP; cap the thread count with `options(fmriAR.max_threads = k)`. You can compute GLS coefficients with a single linear solve:

```{r gls}
Xw <- whitened$X
Yw <- whitened$Y
beta_gls <- qr.solve(Xw, Yw[, 1:5])
beta_gls
```

## Inspecting innovations

The whitened residuals (innovations) should look close to white noise. `acorr_diagnostics()` computes each voxel's autocorrelation, centring each run separately and never pairing frames across a run boundary, and then averages across voxels (`aggregate = "none"` returns the per-voxel matrix instead).

```{r whiteness}
innov_var <- Yw - Xw %*% qr.solve(Xw, Yw)
ac_white <- acorr_diagnostics(innov_var, runs = runs, max_lag = 20)
ac_raw <- acorr_diagnostics(resid, runs = runs, max_lag = 20)

# Mean absolute autocorrelation over lags 1-5, before and after whitening
c(raw = mean(abs(ac_raw$acf[1:5])), whitened = mean(abs(ac_white$acf[1:5])))
```

The raw residuals carry strong low-lag autocorrelation; after whitening it is within the sampling noise band (`ac_white$ci`).

```{r whiteness-plot, fig.cap = "Average autocorrelation across voxels before and after whitening"}
lags <- ac_raw$lags
ylim <- range(c(ac_raw$acf, ac_white$acf, 0))
plot(lags, ac_raw$acf, type = "h", lwd = 2, col = "#1b9e77",
     ylim = ylim, xlab = "Lag", ylab = "Average autocorrelation",
     main = "Mean autocorrelation across voxels")
lines(lags + 0.2, ac_white$acf, type = "h", lwd = 2, col = "#d95f02")
abline(h = c(-1, 1) * ac_white$ci, col = "grey70", lty = 2)
legend("topright",
       legend = c("Raw residuals", "Whitened innovations"),
       col = c("#1b9e77", "#d95f02"),
       lwd = 2, bty = "n")
```

## Censored (scrubbed) frames

Motion-corrupted frames can be passed as `censor`, either as 1-based indices or as a logical mask with one entry per timepoint. Censored frames are dropped when the noise model is estimated: no lag product spans a gap, and the mean is still removed per run. When whitening, the filter restarts (with the exact stationary start) at the first frame after each censored one.

```{r censor}
scrub <- rep(FALSE, n_time)
scrub[c(50:52, 170)] <- TRUE

plan_cens <- fit_noise(resid, runs = runs, p = "auto", censor = scrub)
w_cens <- whiten_apply(plan_cens, X, Y, runs = runs)
w_cens$censor
```

`whiten_apply()` reuses the plan's censor set when it is applied to data of the length it was fitted on, as here; pass `censor` explicitly for other data. The censored rows are still present in the output (filtered from the preceding history), and the result lists them in `censor` so they can be dropped before the final fit:

```{r censor-fit}
keep <- -w_cens$censor
beta_cens <- qr.solve(w_cens$X[keep, ], w_cens$Y[keep, 1:5])
round(beta_cens, 3)
```

## Correcting residual-projection bias

Autocorrelation estimated from GLM residuals is biased towards zero, because the residuals are `M y` for the residual-forming projection `M = I - X (X'X)^-1 X'`, and so carry `M Sigma M` rather than `Sigma`. The bias grows with the number of regressors, and slow drift regressors make it larger still. Our two-column design barely biases anything, so here we add the usual per-run intercepts and low-frequency (DCT) drift terms. Passing the design that produced the residuals undoes the bias:

```{r bias-correction}
dct <- function(L, k) sapply(seq_len(k), function(j) cos(pi * j * (seq_len(L) - 0.5) / L))
drift <- rbind(cbind(dct(120, 4), matrix(0, 120, 4)),
               cbind(matrix(0, 120, 4), dct(120, 4)))
X_full <- cbind(X, run2 = as.numeric(runs == 2), drift)
resid_full <- Y - X_full %*% qr.solve(X_full, Y)

plan_raw <- fit_noise(resid_full, runs = runs, p = 2)
plan_cor <- fit_noise(resid_full, runs = runs, p = 2, design = X_full)
round(rbind(uncorrected = plan_raw$phi[[1]],
            corrected = plan_cor$phi[[1]],
            truth = colMeans(phi_true)), 3)
```

The correction works for every pooling mode with `method = "ar"`, including `pooling = "parcel"`, and with censoring. It requires residuals that really are OLS residuals from `design` (a numerical orthogonality check rejects obvious mismatches). When many datasets share one design, build the correction once with `acvf_bias_matrix()` and pass it as `acvf_correction`. `noise_acvf(resid, design = X)` returns the corrected autocovariance itself.

## Parcel pooling and multiscale shrinkage

When per-voxel residuals are noisy, you can pool information across parcels. The example below constructs synthetic parcels and uses the multiscale PACF-weighted shrinkage.

Under the hood, `fit_noise(..., pooling = "parcel")` estimates each parcel's AR model from the autocovariances of its voxels, pooled across those voxels with run-aware, censor-aware lag products. It does not average the voxels into a parcel-mean series first: that series' autocorrelation is dominated by whatever the voxels share and would not describe the voxel-level noise being whitened. With `parcel_sets`, this is done at each spatial resolution (fine/medium/coarse), and each fine parcel's coefficients are then shrunk toward its parents by combining either partial autocorrelations ("pacf_weighted") or autocovariance functions ("acvf_pooled"). The weights depend on parcel size, the relative dispersion of voxel variances (a unit-free measure, so rescaling the data does not change the result), and the number of runs. Larger or more homogeneous parcels therefore lend more stability to their descendants, and every final per-parcel filter is stationary.

```{r multiscale}
parcels_fine <- rep(1:12, each = n_vox / 12)
parcels_medium <- rep(1:6, each = n_vox / 6)
parcels_coarse <- rep(1:3, each = n_vox / 3)

plan_parcel <- fit_noise(
  resid,
  runs = runs,
  method = "ar",
  p = "auto",
  p_max = 6,
  pooling = "parcel",
  parcels = parcels_fine,
  parcel_sets = list(
    fine = parcels_fine,
    medium = parcels_medium,
    coarse = parcels_coarse
  ),
  multiscale = "pacf_weighted"
)

plan_parcel$order
```

When parcel pooling is requested, `whiten_apply()` returns `X_by`, a list of whitened design matrices per parcel, along with the voxel-wise `Y` innovations. Each `X_by[[pid]]` has the same dimensions as the original `X`; parcels that share a filter share one whitened design.

```{r whiten-parcel}
whitened_parcel <- whiten_apply(plan_parcel, X, Y, runs = runs, parcels = parcels_fine)
length(whitened_parcel$X_by)
```

Parcel coefficients differ from one another, which is the point of parcel pooling: each parcel gets a filter matched to its own noise.

```{r whiten-parcel-check}
# Per-parcel AR coefficients, and confirmation that whitening changed the data
head(sapply(plan_parcel$phi_by_parcel, function(p) round(p, 3))[, 1:4])
max(abs(whitened_parcel$Y - Y))
```

## Fitting ARMA models

Some datasets require ARMA models to capture remaining MA structure. The workflow mirrors the AR case but sets `method = "arma"`.

ARMA models are estimated by the Hannan–Rissanen procedure, pooled over voxels: every voxel is treated as a replicate series sharing the coefficients, so the estimate reflects voxel-level noise rather than the voxel-mean series. All lags stay inside contiguous run/censor segments. `hr_iter` adds refinement iterations, and the plan's `sigma2` is the voxel-scale innovation variance.

```{r arma}
plan_arma <- fit_noise(
  resid,
  runs = runs,
  method = "arma",
  p = 2,
  q = 1,
  hr_iter = 1
)

plan_arma$order
```

Orders can be chosen automatically: `p = "auto"` and `q = "auto"` search `0:p_max` and `0:q_max` by BIC on the pooled regression. Because voxels that share fluctuations carry less independent evidence than their count suggests, the BIC sample size uses the effective number of independent voxels.

```{r arma-auto}
plan_arma_auto <- fit_noise(resid, runs = runs, method = "arma",
                            p = "auto", q = "auto", q_max = 2)
plan_arma_auto$order
```

Here the noise is AR(2), and the search finds no MA component.

ARMA whitening uses the same `whiten_apply()` interface.

```{r arma-whiten}
whitened_arma <- whiten_apply(plan_arma, X, Y, runs = runs)
# The task regressor is 0 for the first 30 TRs in this simulation,
# so showing the first 5 rows hides the effect on that column.
# Instead, inspect rows around the first task onset (t = 31)
# to see how the step regressor is filtered.
whitened_arma$X[28:36, ]
```

```{r arma-design-plot, fig.cap = "Task regressor around onset: raw vs whitened (ARMA)"}
# Visualize how whitening transforms the task step regressor around its first onset
task_raw <- X[, "task"]
task_white <- whitened_arma$X[, "task"]
onset <- which(task_raw > 0)[1]
win <- max(1, onset - 10):min(n_time, onset + 30)
ylim <- range(c(task_raw[win], task_white[win]))
plot(win, task_raw[win], type = "s", lwd = 2, col = "#1b9e77",
     ylim = ylim, xlab = "Time (TR)", ylab = "Regressor value",
     main = "Task regressor: raw vs whitened (ARMA)")
lines(win, task_white[win], lwd = 2, col = "#d95f02")
legend("topleft", legend = c("Raw task", "Whitened task"),
       col = c("#1b9e77", "#d95f02"), lwd = 2, bty = "n")
```

## AFNI-style restricted AR filters

If you need to replicate AFNI's "restricted ARMA" family—real + complex AR roots with optional additive white noise—`fmriAR` exposes a stable wrapper via its compatibility interface: use `compat$afni_restricted_plan()` to construct matching whitening plans. This returns the same `fmriAR_plan` objects used elsewhere while keeping the main public API minimal. Alternatively, you can call the exported top-level helper `afni_restricted_plan()` directly; the compat wrapper delegates to it.

The helper expects the AFNI root parameterisation (a real root `a`, complex pairs defined by radius/angle, and an optional MA(1) refinement) and can operate globally or per parcel. Because the interface mirrors AFNI's internals, it is a convenient bridge when comparing pipelines or validating fmriAR against established AFNI analyses.

```{r afni-plan}
# AFNI-style AR(3): a real root and one complex pair (radius r1, angle t1)
roots <- list(a = 0.6, r1 = 0.7, t1 = pi / 6)

plan_afni <- compat$afni_restricted_plan(
  resid = resid,
  runs = runs,
  parcels = NULL,
  p = 3L,
  roots = roots,
  estimate_ma1 = TRUE
)

plan_afni$order

# Apply whitening just like any other plan
whitened_afni <- whiten_apply(plan_afni, X, Y, runs = runs)
```

As in AFNI, `a`, `r1` and `r2` must be non-negative (values above 0.95 are clamped) and angles are clamped to `[0, pi]`. AFNI models additive white noise through `vrt = s^2 / (s^2 + w^2)`, the share of variance carried by the AR signal. With `estimate_ma1 = FALSE`, a `vrt` below 1 produces the exact ARMA(p, p) equivalent of AR(p) plus white noise, whose autocorrelation is `vrt` times the AR autocorrelation at every non-zero lag. With `estimate_ma1 = TRUE`, an MA(1) term is instead estimated from the AR-filtered voxels, and `vrt` is ignored.

```{r afni-vrt}
plan_afni_vrt <- afni_restricted_plan(resid, runs = runs, p = 3L,
                                      roots = c(roots, vrt = 0.8),
                                      estimate_ma1 = FALSE)
plan_afni_vrt$order
```

Set `estimate_ma1 = FALSE` and leave `vrt` at 1 to obtain the pure AR filter implied by the AFNI roots, or provide a list of root specifications keyed by parcel ID to drive parcel-specific filters that mirror AFNI's voxel grouping.

## Diagnostics and sandwich estimates

The package includes helpers for quick diagnostics and variance estimation:

```{r diagnostics}
# Autocorrelation diagnostics for whitened residuals (innovations), per run
acorr <- acorr_diagnostics(innov_var[, 1:3], runs = runs, max_lag = 5)
round(acorr$acf, 3)

# Standard errors from the whitened fit. "iid" assumes the whitening worked;
# "hac" (Newey-West within runs) stays valid if some autocorrelation remains.
se_iid <- sandwich_from_whitened_resid(whitened$X, whitened$Y[, 1:3])
se_hac <- sandwich_from_whitened_resid(whitened$X, whitened$Y[, 1:3],
                                       type = "hac", runs = runs)
# Rows of $se follow the columns of X; row 2 is the task effect
rbind(iid = se_iid$se[2, ], hac = se_hac$se[2, ])
```

`sandwich_from_whitened_resid()` also accepts rank-deficient designs: non-estimable coefficients get `NA` standard errors, as in `lm()`.

## Next steps

- Explore the test suite in `tests/testthat` for additional usage patterns, including multiscale pooling and ARMA recovery checks.
- Inspect the exported `fmriAR_plan` object with `compat$plan_info(plan_ar)` to see per-run or per-parcel coefficients.
- Combine `whiten_apply()` with GLS modelling packages to build end-to-end prewhitened analyses.

For production usage, consider benchmarking with your acquisition parameters. `tools/validation/accuracy_benchmark.R` in the source repository runs seeded accuracy, calibration and timing scenarios against any installed version of the package.
